---
title: "German_Weimar_ENP"
author: "Frank"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include=FALSE}

  knitr::opts_chunk$set(echo = TRUE)
  
  suppressPackageStartupMessages({
    library(dplyr)
    library(ggplot2)
    library(readr)
    library(here)
    library(knitr)
  })
  
```
  
  ## Germany ENP Analysis
  
  This is my analysis of the Effective Number of Parties in Germany for elections 
  from 1900-1932. I have added insighful analysis of the election formula and some history in Germany.
  
```{r data-import, include = FALSE}
  # --- Load exactly your file and object ---
  load(url("https://www.dropbox.com/scl/fi/kp1lty1chtgzye8gnufl4/clea_lc_20240419.RData?rlkey=phvtoh91a48v9jky1m1120vnw&st=h7lnys7v&dl=1"))
  
   # Loads clea_lc_20240419
  
  
  missing_codes_num  <- c(-990, -992, -994)
  missing_codes_char <- as.character(missing_codes_num)
  
  clea_lc_20240419 <- clea_lc_20240419 %>%
    # numeric columns
    mutate(across(
      where(is.numeric),
      ~ ifelse(. %in% missing_codes_num, NA, .)
    )) %>%
    # character columns
    mutate(across(
      where(is.character),
      ~ ifelse(. %in% missing_codes_char, NA, .)
    )) %>%
    # factor columns
    mutate(across(
      where(is.factor),
      ~ {
        x <- as.character(.)
        x[x %in% missing_codes_char] <- NA_character_
        factor(x)
      }
    ))
  
  # --- Scope: German, 1900–1932 ---
  #
countries <- c("Russian Federation")
russia <- clea_lc_20240419 %>%
    filter(ctr_n %in% countries, yr >= 2000, yr <= 2025)
  
```
  
  ## Calculating the ENP for 1903 for Constituency 1
  
  This replicates what we did for Data Lab 1
  
  
```{r enp_2013_cst1, echo=FALSE}
russia_2016 <- russia %>%
    filter(yr == 2016) %>%
    group_by(cst, cst_n) %>%
    summarize(enp_v = 1/sum(pvs1^2),
              enp_s = 1/sum((seat/mag)^2),
              yr = first(yr),
              mag = first(mag),
              cst_n = first(cst_n)) %>%
    ungroup()
  
kable(russia_2016, caption = "ENP for Germany, 1903, Constituency 1")
  
```
  
## Appears to be no seats for Germany in 1903.

Perhaps there are no seat values?

```{r check_seats, echo=FALSE}

germany %>%
  select(seat,mag) %>%
    glimpse()

# Looks like all NA values. Need to just use EPN(v)

```

  
```{r enp_v_all, echo=FALSE}

germany_1903_alldistricts <- germany %>%
    filter(yr == 1903) %>%
    group_by(cst, cst_n) %>%
    summarize(enp_v = 1/sum(pvs1^2),
#              enp_s = 1/sum((seat/mag)^2),
              yr = first(yr),
              mag = first(mag),
              cst_n = first(cst_n)) %>%
    ungroup()
  
kable(germany_1903_alldistricts, caption = "ENP for Germany, 1903, All Constituencies")
  
# Histogram of ENP(v)
max_v <- max(germany_1903_alldistricts$enp_v, na.rm = TRUE)

ggplot(germany_1903_alldistricts, aes(x = enp_v)) +
  geom_histogram(binwidth = 0.25, boundary = 1, closed = "left") +
  # (Optional) add a smooth density curve on top
  geom_density(aes(y = after_stat(density)), linewidth = 1) +
  scale_x_continuous(limits = c(1, max_v), breaks = scales::pretty_breaks(n = 6)) +
  labs(
    title = "Distribution of ENP (Votes), Germany 1903",
    x = "ENP (Votes)",
    y = "Number of Constituencies"
  ) +
  theme_minimal()
  
```
  
## Calculating ENP for all years from 1900-1932  
  
  
```{r enp_all, echo=FALSE}

germany_1900_1932 <- germany %>%
    filter(yr >= 1900 & yr <= 1932) %>%
    group_by(yr, cst, cst_n) %>%
    summarize(enp_v = 1/sum(pvs1^2),
#              enp_s = 1/sum((seat/mag)^2),
              yr = first(yr),
              mag = first(mag),
              cst_n = first(cst_n)) %>%
    ungroup()
  
ggplot(germany_1900_1932, aes(x = enp_v)) +
  geom_histogram(binwidth = 0.25, boundary = 1, closed = "left") +
  # (Optional) add a smooth density curve on top
  geom_density(aes(y = after_stat(density)), linewidth = 1) +
  facet_wrap(~ yr, scales = "free") +
  scale_x_continuous(limits = c(1, 7), breaks = scales::pretty_breaks(n = 6)) +
  labs(
    title = "Distribution of ENP (Votes), Germany 1903",
    x = "ENP (Votes)",
    y = "Number of Constituencies"
  ) +
  theme_minimal()
  
```
  
