---
title: "Italy_ENP"
author: "Dr. Paul Gronke"
date: "`r Sys.Date()`"
output: pdf_document
---
  
```{r setup, include=FALSE}
  # --------------------------------------------
  # ENP from CLEA constituency data (2000-2025)
  # Countries: Denmark and Italy
  # Object in .RData: clea_lc_20240419
  # Uses: CTR, YR, CST, MAG, PTY, PV1 (votes, round 1), SEAT
  # --------------------------------------------
  
  knitr::opts_chunk$set(echo = TRUE)
  
  suppressPackageStartupMessages({
    library(dplyr)
    library(ggplot2)
    library(readr)
    library(here)
    library(knitr)
  })
  
```
  
  ## Italy ENP Analysis
  
  This is my analysis of the Effective Number of Parties in Italy for elections 
  from 2000 - 2013. I have added insighful analysis of the election formula in Italy.
  
```{r data-import, include = FALSE}
  # --- Load exactly your file and object ---
  load(url("https://www.dropbox.com/scl/fi/kp1lty1chtgzye8gnufl4/clea_lc_20240419.RData?rlkey=phvtoh91a48v9jky1m1120vnw&st=h7lnys7v&dl=1"))
  
   # Loads clea_lc_20240419
  
  
  missing_codes_num  <- c(-990, -992, -994)
  missing_codes_char <- as.character(missing_codes_num)
  
  clea_lc_20240419 <- clea_lc_20240419 %>%
    # numeric columns
    mutate(across(
      where(is.numeric),
      ~ ifelse(. %in% missing_codes_num, NA, .)
    )) %>%
    # character columns
    mutate(across(
      where(is.character),
      ~ ifelse(. %in% missing_codes_char, NA, .)
    )) %>%
    # factor columns
    mutate(across(
      where(is.factor),
      ~ {
        x <- as.character(.)
        x[x %in% missing_codes_char] <- NA_character_
        factor(x)
      }
    ))
  
  # --- Scope: Italy, 1980–2005 ---
  #
  countries <- c("Italy")
  italy <- clea_lc_20240419 %>%
    filter(ctr_n %in% countries, yr >= 2000, yr <= 2025)
  
```
  
  ## Calculating the ENP for 2013 for Constituency 1
  
  This replicates what we did for Data Lab 2
  
  
```{r enp_2013_cst1, echo=FALSE}
  italy_2013_district1 <- italy %>%
    filter(yr == 2013 & cst == 1) %>%
    group_by(cst, cst_n) %>%
    summarize(enp_v = 1/sum(pvs1^2),
              enp_s = 1/sum((seat/mag)^2),
              yr = first(yr),
              mag = first(mag),
              cst_n = first(cst_n)) %>%
    ungroup()
  
  kable(italy_2013_district1, caption = "ENP for Italy, 2013, Constituency 1")
  
```
  
  ## Calculating the ENP for 2013 for All Constituencies
  
  This replicates what we did for Data Lab 2 but for all constituencies in 2013
  
  
```{r enp_2013_all, echo=FALSE}
  italy_2000_2025 <- italy %>%
    filter(yr > 2000) %>%
    group_by(yr, cst, cst_n) %>%
    summarize(enp_v = 1/sum(pvs1^2),
              enp_s = 1/sum((seat/mag)^2),
              yr = first(yr),
              mag = first(mag),
              cst_n = first(cst_n)) %>%
    ungroup()
  
  kable(italy_2000_2025, caption = "ENP for Italy, All Years, All Constituencies")
  
  ggplot(italy_2000_2025, aes(x = enp_v, y = enp_s)) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    # Use LOESS instead of linear fit
    geom_smooth(method = "loess", formula = y ~ x, se = FALSE, color = "blue") +

    # Axes from 1 up to the overall max
    scale_x_continuous(limits = c(1, 6)) +
    scale_y_continuous(limits = c(1, 6)) +
    labs(title = "ENP Votes vs ENP Seats for Italy, 2013",
         x = "ENP Votes",
         y = "ENP Seats") +
    theme_minimal()
  
```
  
## Calculating ENP for all years from 2000-2025  
  
  
```{r enp_all, echo=FALSE}
italy_2013_alldistricts <- italy %>%
    filter(yr == 2013) %>%
    group_by(cst, cst_n) %>%
    summarize(enp_v = 1/sum(pvs1^2),
              enp_s = 1/sum((seat/mag)^2),
              yr = first(yr),
              mag = first(mag),
              cst_n = first(cst_n)) %>%
    ungroup()
  
  kable(italy_2013_alldistricts, caption = "ENP for Italy, 2013, All Constituencies")
  
ggplot(italy_2013_alldistricts, aes(x = enp_v, y = enp_s)) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    # Use LOESS instead of linear fit
    geom_smooth(method = "loess", formula = y ~ x, se = FALSE, color = "blue") +
    facet_wrap(~ yr) +
    # Axes from 1 up to the overall max
    scale_x_continuous(limits = c(1, 6)) +
    scale_y_continuous(limits = c(1, 6)) +
    labs(title = "ENP Votes vs ENP Seats for Italy, 2013",
         x = "ENP Votes",
         y = "ENP Seats") +
    theme_minimal()
  
```

## Nationwide ENP in Italy 

```{r nationwide_enp, echo=FALSE}
nat_party_totals <- italy %>%
  group_by(ctr, ctr_n, yr, pty, .drop = TRUE) %>%
  summarize(
    votes = sum(pv1, na.rm = TRUE),
    seats = sum(seat, na.rm = TRUE),
    ctr_n = first(ctr_n),
    ctr = first(ctr),
    pty = first(pty),
    pty_n = first(pty_n),
    .groups = "drop"
  )

nat_totals <- italy %>%
  group_by(ctr_n, yr, .drop = TRUE) %>%
  summarize(
    totalvotes = sum(pv1, na.rm = TRUE),
    totalseats = sum(seat, na.rm = TRUE),
    ctr_n = first(ctr_n),
    .groups = "drop"
  )

nat_party_totals <- nat_party_totals %>%
  left_join(nat_totals, by = c("ctr_n", "yr")) %>%
  mutate(vote_share = votes / totalvotes,
         seat_share = seats / totalseats)

nat_enp <- nat_party_totals %>%
    group_by(ctr_n, yr) %>%
    summarize(enp_v = 1/sum(vote_share^2),
              enp_s = 1/sum(seat_share^2),
              yr = first(yr),
              ctr_n = first(ctr_n)
              ) %>%
    ungroup()

ggplot(nat_enp, aes(x = yr)) +
  geom_line(aes(y = enp_v, color = "Votes"), linewidth = 1.1) +
  geom_point(aes(y = enp_v, color = "Votes"), size = 2) +
  geom_line(aes(y = enp_s, color = "Seats"), linewidth = 1.1) +
  geom_point(aes(y = enp_s, color = "Seats"), size = 2) +
  facet_wrap(~ ctr_n, scales = "free_y") +
  scale_color_manual(values = c("Votes" = "darkgreen", "Seats" = "darkred")) +
  labs(
    title = "Nationwide Effective Number of Italian Parties (ENP), 2000-2025",
    x = "Election Year",
    y = "ENP"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```
  
