---
title: "Assignment_3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

## Step 1: Read in the SPAE 

1. The SPAE Report is available here https://electionlab.mit.edu/research/projects/survey-performance-american-elections
2. The SPAE Data are avaiable at Datavers https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/GWBAP5

I encourage you to learn how to download data yourself from dataverse, but for the purposes
of this lab, I have stored a Stata and SPSS version of the SPAE in the Assignment folder. 

To read the SPAE from either Stata or SPSS, you need to have the haven package. We will
also load the tidyverse package. 

Make sure you make the working directory so that it is "looking" in the "Assignment_3_SPAE" folder. 

```{r read_data}
library(tidyverse)
library(haven)
library(janitor) # creates nice tables using the tabyl command and supports piping

# Read the SPAE data from Stata
spae_stata <- read_dta("mitu0051_output.dta")
spae_spss <- read_sav("MITU0051_OUTPUT.sav")

```

## Step 2: Explore the SPAE

Browse the data. See if there are any key differences between the SPSS and Stata versions. 

```{r explore_data}
glimpse(spae_stata)
glimpse(spae_spss)

summary(spae_stata)
summary(spae_spss)

table(spae_spss$Q1)
table(spae_spss$pid3)

spae_spss %>%
 tabyl(Q1)

spae_spss %>%
 tabyl(pid3)

```

```{r sjlabelled_tables}
library(sjlabelled)

table(as_label(spae_spss$Q1))
table(as_label(spae_spss$pid3))

# spae_spss %>%
#   tabyl(as_label(spae_spss$Q1))
# 
# spae_spss %>%
#   tabyl(as_label(spae_spss$pid3))

```

## Step 4: Learn how to use survey weights in R

R has a package called "survey" that allows you to use survey weights. There is also a
package called "srcvyr" that allows you to use the "dplyr" syntax with survey weights. 

Finally, there is a package called "gtsummary" that recognizes "survey weighted objects", 
which means you can make nice tables with survey weighted data.

You should install these if you do not already have them.

Documentation for using survey data in R is provided at https://tidy-survey-r.github.io/tidy-survey-book/


```{r survey_weighted_tables}
library(survey)
library(srvyr)
library(gtsummary)

# Convert all   columns which are "labeled" as factors since these will display
# better in tables in R. 

spae_spss <- spae_spss %>%
  mutate(across(
    where(haven::is.labelled),
    ~ haven::as_factor(.x, levels = "labels")
  ))

# This alternative method converts only specific columns to factors

# spae_spss <- spae_spss %>%
#   mutate(
#     Q1   = haven::as_factor(Q1,   levels = "labels"),
#     pid3 = haven::as_factor(pid3, levels = "labels")
#   )

# Create a survey design object


spae_spss_design <- spae_spss %>%
  as_survey_design(weights = weight_stacked)

# Create a survey weighted table using gtsummary

# spae_spss_design %>%
#   tabyl(., Q1) 
#   
spae_spss_design %>%
  select(Q1) %>%
  tbl_svysummary(.)

spae_spss_design %>%
  select(pid3) %>%
  tbl_svysummary(.)

# Now show how you can do this in one pipe without creating a separate object

spae_spss %>% as_survey_design(., weights = weight_stacked) %>%
  tbl_svysummary(include =Q1) %>%
  modify_header(label ~ "**PID3**") %>%
  modify_caption("**Weighted distribution of PID3**")


```

























